---
title: "APPAC: correcting GC peak areas for atmospheric pressure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{APPAC: correcting GC peak areas for atmospheric pressure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2, patchwork}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 5, dpi = 96,
  message = FALSE, warning = FALSE
)
## the plotting functions need the suggested ggplot2 / patchwork; skip those
## chunks gracefully if they are not installed.
have_plots <- requireNamespace("ggplot2", quietly = TRUE) &&
              requireNamespace("patchwork", quietly = TRUE)
```

**APPAC** (*Atmospheric Pressure Peak Area Correction*) corrects gas-chromatography
peak areas for the influence of ambient **air pressure** on standard detectors open
to the atmosphere — chiefly the flame-ionization detector (FID). This vignette first
shows the everyday **usage** on the bundled `PLOT_FID` data, then explains the
**method**: how a single common pressure sensitivity, per-cylinder drift and
measurement noise are separated by a principal-component decomposition.

# Usage

APPAC ships the `PLOT_FID` example dataset — six control cylinders, five hydrocarbon
peaks and roughly five years of real FID injections with expert-annotated peak
integration. The plots use the suggested `ggplot2` / `patchwork`.

```{r packages}
library(appac)
```

## Load the data

`PLOT_FID` is long-format (one row per injection × peak). `check_cols()` maps the raw
column names to the canonical ones and cleans the sample/peak names. The reference
pressure `P_ref` is the mid-range of the observed air pressure.

```{r load}
acn <- list(sample_col = "sample.name", peak_col = "peak.name",
            date_col = "injection.date", pressure_col = "air.pressure",
            area_col = "raw.area")
data  <- check_cols(PLOT_FID, acn)
ap    <- as.numeric(data[, "Air_Pressure"])
P_ref <- mean(range(ap, na.rm = TRUE))
cat(sprintf("%d injections, %d cylinders, %d peaks;  P_ref = %.1f hPa\n",
            nrow(data) / length(unique(data$Peak_Name)),
            length(unique(data$Sample_Name)),
            length(unique(data$Peak_Name)), P_ref))
```

## Fit the correction

Two passes: `appac()` estimates the common pressure sensitivity **kappa** and the
drift model; `debias_ct()` then finds the centre that minimises the chi-square of the
residuals (the bias–variance trade-off), and a second `appac()` runs with that
de-biased centre. *(`npt = 7` keeps the vignette quick; production uses `npt = 21`.)*

```{r fit}
fit1 <- appac(data = data, P_ref = P_ref)
ct   <- debias_ct(fit1, data = data, P_ref = P_ref, npt = 7, quiet = TRUE)
fit  <- appac(data = data, ct = ct, P_ref = P_ref)
cat(sprintf("kappa = %.3e  (per hPa)\n", unlist(fit@correction@coefficients)))
```

## Inspect the fit

`plot_area_pressure_fit()` shows exactly what enters the kappa estimate: the
reference-scaled correlated area for every (cylinder, peak), binned by pressure
deviation. All series share the single common slope kappa.

```{r kappa-fit, eval = have_plots}
plot_area_pressure_fit(fit)
```

Raw areas scatter with the day-to-day pressure swings; the corrected areas collapse
onto the reference. Detected episode breakpoints are marked.

```{r area-date, eval = have_plots}
plot_area_date(fit, sample = 1, peak = "n.C4H10", show_changepoints = TRUE)
```

## Run-to-run scatter and diagnostics

The headline metric — relative standard deviation of the areas before and after
correction (cylinder 1):

```{r rsd}
rsd <- function(x) sd(x) / mean(x) * 100
raw <- fit@samples[[1]]$raw.area
cor <- fit@samples[[1]]$corrected.area
data.frame(peak              = colnames(raw),
           RSD_raw_pct       = round(apply(raw, 2, rsd), 3),
           RSD_corrected_pct = round(apply(cor, 2, rsd), 3))
```

Reduced chi-square ≈ 1 means the corrected areas are down to the short-term
measurement-noise floor:

```{r gof}
goodness_of_fit(fit)[[1]]
```

`plot_residuals()` panels the residual diagnostics — a (leptokurtic, heavy-tailed)
histogram with a fitted normal, the normal Q–Q, and residual vs. date / vs. pressure.
Both scatter panels are flat, confirming the structured pressure and drift artifacts
are gone and only the measurement noise remains.

```{r residuals, fig.width = 10, fig.height = 8, eval = have_plots}
plot_residuals(fit, sample = 1, peak = "n.C4H10")
```

A compact summary of the fitted object (`print()` adds the per-sample goodness of
fit):

```{r show}
fit
```

# Method: the decomposition

## The physical effect and the forward model

Boček, Novák & Janák (1969) showed experimentally that the FID's ionization
efficiency is strongly pressure-dependent: normal atmospheric swings already change
the sensitivity by up to about 5 % across the naturally occurring range of ambient
pressure. In routine
data this pressure signal is **masked by slow instrument drift**, so it must be
*separated* from the drift before it can be corrected.

APPAC models the raw area of peak \(j\) in cylinder \(i\) at injection \(t\) as a
multiplicative combination of a reference area, a pressure factor, a slow
drift / daily factor and noise:

\[
A_{ijt} \;=\; \mu_{ij}\,\bigl(1 + \kappa\,(P_t - P_\mathrm{ref})\bigr)\,
              (1 + \delta_{it})\,(1 + \varepsilon_{ijt}),
\]

where \(\mu_{ij}\) is the per-cylinder, per-peak reference (centre), \(\kappa\) the
**common** pressure sensitivity shared by all peaks and cylinders, \(P_t\) the ambient
pressure, \(\delta_{it}\) a slow drift plus a daily factor, and
\(\varepsilon_{ijt}\) heavy-tailed measurement noise.

## Separating the components by PCA

Per cylinder, each peak is standardised and a one-component principal-component
analysis is run on the (dates × peaks) matrix. The components carry distinct physics:

* **PC1 — correlated component:** the dominant common mode across peaks — the
  pressure response together with the smooth drift.
* **PC2 — uncorrelated component:** the second mode, carrying the per-cylinder bias
  and *abrupt* level shifts (the change-point signal).
* **PC3 and higher — noise:** the per-peak, injection-to-injection measurement noise.

Because the noise is multiplicative — the per-peak standard deviation grows with the
level, \(\mathrm{sd} \approx a\,\mu + b\,\mu^2\) — a large peak would otherwise
dominate the PCA. Each peak is therefore referenced to its centre and
**dispersion-normalised**: divided by its modelled dispersion so that every peak
enters the PCA with comparable variance. This is the variance-stabilising
(diagonal-scaling) half of a *whitening* transform; the PCA then supplies the
decorrelation, so standardise-then-PCA together sphere the per-peak signals. The
whitening is *across peaks* (amplitude), not *in time*: the injection-to-injection
noise itself stays reddened (brown, AR(1)-like), as the `Synth_data` fixture
deliberately illustrates.

## Estimating the common kappa

Kappa is estimated on a **drift-reduced** signal: a robust loess low-pass removes the
slow drift on a *copy* of the areas, so \(\kappa\) is not confounded with the drift.
The correlated component is then regressed on the pressure deviation with a
heavy-tail-robust generalized linear model (`robustbase::glmrob`), which down-weights
the tails instead of being dragged by them; the single common slope yields
\(\kappa = 1/\sum_k \beta_k\). Robust regression is the appropriate tool here because
the corrected-area distribution is markedly leptokurtic.

## Drift and the daily factor (NA-tolerant imputation)

The uncorrelated (PC2) daily factors are combined **across cylinders** to estimate the
instrument-wide drift and bias. Because the cylinders are not all measured on the same
days, the cross-cylinder daily-factor matrix has gaps. APPAC fills them by
reconstruction rather than dropping incomplete days: a base-R NA-tolerant PCA
(unit-variance scaling, gaps seeded at the mean, then an iterated rank-\(k\) SVD
reconstruction — the svdImpute / EM scheme of probabilistic PCA; Tipping & Bishop
1999; Roweis 1998; Dempster, Laird & Rubin 1977). The first two components give the
common trend and per-cylinder bias; a parallel decomposition of the correlated part
gives the common daily factor. The assembled multiplier (kappa pressure factor +
drift + daily factor) is then divided out of the raw areas.

## Refining the centre: chi-square minimisation

The correction references each peak to a centre \(\mu_{ij}\); the whole-series mean is
a convenient first estimate but a biased one — it is itself pulled by the residual
structure the correction has not yet removed. `debias_ct()` refines it by
**minimising the residual chi-square**. For a small grid of scale factors \(c\)
around 1, it re-runs `appac()` with the centre scaled to \(c\,\mu_{ij}\) and records,
per peak, the chi-square of the corrected-area residuals about that centre. That
statistic is very nearly quadratic in \(c\), so its minimum is the closed-form
parabola vertex

\[
\chi^2(c) \;\approx\; b_0 + b_1 c + b_2 c^2 \ (b_2 > 0),
\qquad
c^{*} \;=\; -\,\frac{b_1}{2\,b_2},
\]

with \(b_0, b_1, b_2\) fitted by ordinary least squares over the sweep. The de-biased
centre \(c^{*}\mu_{ij}\) trades a little variance for reduced bias and is fed back
into a second `appac()` pass. (The closed form replaces a root-find on the
derivative, which failed whenever the optimum fell outside the sweep — precisely the
large-bias case the refinement exists for.)

## Change points: episode and precision breaks

The PC2 drift signal carries **real structural breaks** — cylinder changes, detector
maintenance, column trims. `get_changepoints()` detects these episode boundaries with
a deterministic structural-break model (an OLS-MOSUM fluctuation test gating
BIC-optimal breakpoint dating from **strucchange**; Zeileis et al. 2002). Its
second-moment counterpart, `get_variance_changepoints()`, detects changes in
measurement **precision** (variance) on the noise-energy signal — a distinct
instrument event from a level shift.

```{r changepoints}
get_changepoints(fit@samples)            # episode level breakpoints
```

Because these breaks are what separates one calibration regime from the next, they are
central to the correction rather than an afterthought — and, unlike the
single-calibration-period pressure compensation of Ayers & Clardy (1985), let APPAC
correct a multi-year record without per-period recalibration.

# References

* Boček P., Novák J., Janák J. (1969). Effect of pressure on the performance of the
  flame ionization detector. *J. Chromatogr.* **43**, 431–436.
  <doi:10.1016/S0021-9673(00)99223-9>
* Ayers B. O., Clardy D. E. (1985). *Barometric pressure correction for a chromatograph.*
  US Patent 4,512,181.
* Tipping M. E., Bishop C. M. (1999). Probabilistic principal component analysis.
  *J. R. Stat. Soc. B* **61**(3), 611–622.
* Roweis S. (1998). EM algorithms for PCA and SPCA. *Adv. Neural Inf. Process. Syst.* **10**.
* Dempster A. P., Laird N. M., Rubin D. B. (1977). Maximum likelihood from incomplete
  data via the EM algorithm. *J. R. Stat. Soc. B* **39**(1), 1–38.
* Zeileis A., Leisch F., Hornik K., Kleiber C. (2002). strucchange: An R package for
  testing for structural change in linear regression models. *J. Stat. Softw.*
  **7**(2). <doi:10.18637/jss.v007.i02>
